---
title: "EPIC package"
author: "Julien Racle and David Gfeller"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{EPIC package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Description
Package implementing EPIC method to estimate the proportion of immune and cancer
cells from bulk gene expression data. It is based on reference gene expression
profiles for the main immune cell types and it predicts the proportion of these
cells and of the remaining "other cells" that are the remaining cells (cancer,
stromal, endothelial) for which no reference profile is given.

## Usage
The main function in this package is `EPIC`. It needs as input a matrix of the raw gene expression counts from the samples for which to estimate cell proportions. One can also define the reference cells to use
```{r, eval = FALSE}
out <- EPIC(bulk = bulkSamplesMatrix)
out <- EPIC(bulk = bulkSamplesMatrix, reference = referenceCellsList)
```

`out` is a list containing the various mRNA and cell fractions in each samples as well as some *data.frame* of the goodness of fit.

Values of mRNA per cell and signature genes to use can also be changed:
```{r, eval = FALSE}
out <- EPIC(bulk = bulkSamplesMatrix, reference = referenceCellsList, mRNA_cell = mRNA_cell_vector, sigGenes = sigGenes_vector)
out <- EPIC(bulk = bulkSamplesMatrix, reference = referenceCellsList, mRNA_cell_sub = mRNA_cell_sub_vector)
```


## Installation
```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("GfellerLab/EPIC")
```

